Interisland movements of woodland caribou in Newfoundland

### 2020-02-19 paper/manuscript.Rmd
monster reflow

## Materials and methods

### Study area

Our study area was the Fogo Island archipelago, Newfoundland, Canada (49°N, 54°W). Fogo Island is a small (237.71 km2) island off the northeastern coast of Newfoundland and has humid climate with year-round precipitation. While Fogo Island is the largest island in the area, the archipelago includes at least three larger islands: Western Indian (77.6 km2), Eastern Indian (38.7 km2) and Change (XX km2) Islands, as well as numerous smaller islands, including Blundon’s (1.18 km2), North Long (1.01 km2), South Long (0.48 km2), Kate’s (1.64 km2), and Brother’s (1.59 km2) Islands (see Figure 1). All islands in the Fogo Island archipelago consist largely of coniferous and mixed forests of balsam fir (Abies balsamea), black spruce (Picea mariana), and white birch (Betula payrifera) as well as bogs, lakes, lichen and rocky barrens.

@@

### Caribou observation and swimming

@@

### Statistical analysis
We were interested in determining what factors might influence how long an individual remains on the terminal island. We therefore used linear regressions to test how the Julian date of arrival to the island, size of the terminal island, and the paired identity of islands (i.e., Fogo to Perry or Perry to Eastern Indian) influenced the duration of time spent on the terminal island. All statistical analyses were conducted in R.

## Discussion

@@

### Forage limitation and density-dependence
Forage limitation and over-grazing is a major issue for caribou populations (Schaefer et al. 2016; Zamin et al. 2017). Reduced access to foraging opportunities can result in higher conspecific competition for resources and lowered reproductive success for adult females (Schaefer et al. 2016). For caribou in the Fogo Island archipelago, the role of forage limitation in animal habitat selection patterns and movement remains unknown, but it is possible that forage limitation and increased conspecific competition influences swimming between islands.

@@

### Predator avoidance
Predator avoidance could contribute to swimming behaviour of caribou in our system as well as others (Miller 2002). Unlike elsewhere in the range of caribou, wolves are extirpated from Newfoundland, including Fogo Island. In Newfoundland, coyotes are the primary predator of caribou (Bastille-Rousseau et al. 2016a). While predation by coyotes on adult female caribou is possible, it is relatively rare on Fogo Island (Webber & Vander Wal unpublished data). Therefore, any effect of predation on swimming behaviour is likely related to fine-scale encounters between coyotes and caribou which result in caribou swimming as a flight response. Alternatively, during calving, females may swim to small islands to give birth (Bergerud & Page 1987) or swim to islands with a calf at heel to avoid predation. Swimming with a calf at heel has been observed for caribou in freshwater lakes on Fogo Island (Webber & Vander Wal unpublished data), so we anticipate it may also be possible in the ocean. We suggest future studies assess encounter rates among caribou and coyotes to determine what proportion of encounters result in swimming behaviour.

@@

## Acknowledgements
We thank M. Laforge, M. Bonar, C. Hart, and R. Huang for help in the field. Logistical support was provided L. Bixby. We also thank members of the Newfoundland Wildlife Division including S. Moores, B. Adams, K. Lewis, C. Doucet, W. Barney, F. Norman, R. Otto, J. Neville, P. Saunders, T. Porter, P. Tremblett, S. Gullage, T. Hodder, D. Jennings, J. McGinn, M. Frampton, P. McClement, and S. Mahoney for data collection, management, and logistics. We thank T. Bergerud and S. Mahoney for their vision in initiating much of the work on caribou in Newfoundland. Funding for this study was provided by a Vanier Canada Graduate Scholarship to QMRW and a NSERC Discovery Grant to EVW.

>>>>

The Fogo Island archipelago is located in Newfoundland, Canada (49°N, 54°W), and
comprises approximately ten islands. While Fogo Island is the largest island
(237.71 km^2), the archipelago includes at least three larger islands: Western
Indian (77.6 km2), Eastern Indian (38.7 km2) and Change (XX km2) Islands, as
well as numerous smaller islands, including Blundon’s (1.18 km2), North Long
(1.01 km2), South Long (0.48 km2), Kate’s (1.64 km2), and Brother’s (1.59 km2)
Islands (see Figure 1). All islands in the Fogo Island archipelago consist
largely of coniferous and mixed forests of balsam fir (Abies balsamea), black
spruce (Picea mariana), and white birch (Betula payrifera) as well as bogs,
lakes, lichen and rocky barrens.

@@

@@

@@

Forage limitation and over-grazing is a major issue for caribou populations
(Schaefer et al. 2016; Zamin et al. 2017). Reduced access to foraging
opportunities can result in higher conspecific competition for resources and
lowered reproductive success for adult females (Schaefer et al. 2016). For
caribou in the Fogo Island archipelago, the role of forage limitation in animal
habitat selection patterns and movement remains unknown, but it is possible that
forage limitation and increased conspecific competition influences swimming
between islands.

@@

Predator avoidance could contribute to swimming behaviour of caribou in our
system as well as others (Miller 2002). Unlike elsewhere in the range of
caribou, wolves are extirpated from Newfoundland, including Fogo Island. In
Newfoundland, coyotes are the primary predator of caribou (Bastille-Rousseau et
al. 2016a). While predation by coyotes on adult female caribou is possible, it
is relatively rare on Fogo Island (Webber & Vander Wal unpublished data).
Therefore, any effect of predation on swimming behaviour is likely related to
fine-scale encounters between coyotes and caribou which result in caribou
swimming as a flight response. Alternatively, during calving, females may swim
to small islands to give birth (Bergerud & Page 1987) or swim to islands with a
calf at heel to avoid predation. Swimming with a calf at heel has been observed
for caribou in freshwater lakes on Fogo Island (Webber & Vander Wal unpublished
data), so we anticipate it may also be possible in the ocean. We suggest future
studies assess encounter rates among caribou and coyotes to determine what
proportion of encounters result in swimming behaviour.

@@

## Acknowledgements 
We thank M. Laforge, M. Bonar, C. Hart, and R. Huang for help in the field.
Logistical support was provided L. Bixby. We also thank members of the
Newfoundland Wildlife Division including S. Moores, B. Adams, K. Lewis, C.
Doucet, W. Barney, F. Norman, R. Otto, J. Neville, P. Saunders, T. Porter, P.
Tremblett, S. Gullage, T. Hodder, D. Jennings, J. McGinn, M. Frampton, P.
McClement, and S. Mahoney for data collection, management, and logistics. We
thank T. Bergerud and S. Mahoney for their vision in initiating much of the work
on caribou in Newfoundland. Funding for this study was provided by a National
Sciences and Engineering Research Council (NSERC) Vanier Canada Graduate
Scholarship to QMRW, NSERC Canda Graduate Scholarship to JGH, and a NSERC
Discovery Grant to EVW.
